Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-12 paper/paper.tex
small revision on section 5 and 6

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may be interested in counting the independent sets of large sizes close to the MIS size.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Eq{eq:polynomial}:
\begin{equation}
    \begin{split}
    a \oplus b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
As will be shown bellow, this algebra can also be extended to enumerate suboptimal solutions.

@@

This sets algebra can serve as the coefficients in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
For example, to enumerate only the MISs, with the tropical algebra, we define $s(k) = (s_{k}, k)$, where the first element follows the algebra in \Eq{eq:set} and the second element follows the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\begin{split}
    s(k) \oplus s(j) &= \begin{cases}
        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
        \left(s_j, \max(k,j) \right), & k < j \\
        \left(s_k, \max(k,j) \right), & k > j
    \end{cases}, \\
    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
    \mymathbb{0} &= (\{ \}, -\infty)\\
    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
\end{split}
\end{equation}
%\begin{equation}
%\begin{split}
%    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
%        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
%        s_y\infty^{\max(x,y)}, & x < y\\
%        s_x\infty^{\max(x,y)}, & x > y
%    \end{cases},\\
%    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
%    \mymathbb{0} &= \{\}\infty^{-\infty},\\
%    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0},
%\end{split}
%\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \left({\{\boldsymbol{e}_{i}\}, 1} \right)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

If one is interested in obtaining only a single MIS configuration, one can just keep a single configuration in the intermediate computations to save the computational effort.
Here is a new algebra defined on the bit strings, replacing the sets of bit strings in \Eq{eq:set}, 
%We leave this as an exercise for readers.
%
%\iffalse
%\begin{equation}
%\begin{split}
%    \sigma_x\infty^x \oplus \sigma_y\infty^y &= \begin{cases}
%        {\rm select}(\sigma_x, \sigma_y)\infty^{\max(x,y)}, & x = y\\
%        \sigma_y\infty^{\max(x,y)}, & x < y\\
%        \sigma_x\infty^{\max(x,y)}, & x > y
%    \end{cases},\\
%    \sigma_x\infty^x \odot \sigma_y\infty^y &= (\sigma_x \lor^\circ \sigma_y)\infty^{x+y},\\
%    \mymathbb{0} &= 1^{\otimes n}\infty^{-\infty},\\
%    \mymathbb{1} &= 0^{\otimes n}\infty^{0},
%\end{split}
%\end{equation}
%\fi
\begin{equation}
\begin{split}
    \sigma \oplus \tau &= {\rm select}(\sigma, \tau), \\
    \sigma \odot \tau &= (\sigma\lor^\circ \tau),\\
    \mymathbb{0} &= 1^{\otimes |V|}, \\
    \mymathbb{1} &= 0^{\otimes |V|},
\end{split}\label{eq:singleconfig}
\end{equation}
where the \texttt{select} function picks one of $\sigma$ and $\tau$ by some criteria to make the algebra commutative and associative, e.g. by picking one with a smaller integer value.
%In practice, we can just pick randomly from them, in which case the program will output one of the MIS configurations randomly.
\red{Can we also enumerate all maximal independent sets by combining with \Eq{eq:maximal} or enumerate suboptimal solutions by combining with the truncated polynomial?}
\blue{now?}

Although the above discussion is in the context of the the MIS problem, we can also combine the above algebras with the maximal independence network in \Eq{eq:maximal} to enumerate or sample maximal independent sets,
as well as other problems in \App{app:otherproblems} for configuration enumeration and sampling.

\subsection{Bounding the MIS enumeration space}
When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we use $\alpha(G)$ to bound the searching space.
First, we compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra, \Eq{eq:countingtropicalset}, for obtaining all configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagation of the gradients.
To derive the back-propagation rule for tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ being $\overline C$, the back-propagation rule for gradient masks reads
\begin{equation}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations.

>>>>

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may be interested in counting the independent sets of large sizes close to the MIS size.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Eq{eq:polynomial}:
\begin{equation}
    \begin{split}
    a \oplus b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
As will be shown below, this algebra can also be extended to enumerate those large-size independent sets.

@@

This set algebra can serve as the first element in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
For example, to enumerate only the MISs, with the tropical algebra, we define $s(k) = (s_{k}, k)$, where the first element follows the algebra in \Eq{eq:set} and the second element follows the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\begin{split}
    s(k) \oplus s(j) &= \begin{cases}
        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
        \left(s_j, \max(k,j) \right), & k < j \\
        \left(s_k, \max(k,j) \right), & k > j
    \end{cases}, \\
    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
    \mymathbb{0} &= (\{ \}, -\infty)\\
    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
\end{split}
\end{equation}
%\begin{equation}
%\begin{split}
%    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
%        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
%        s_y\infty^{\max(x,y)}, & x < y\\
%        s_x\infty^{\max(x,y)}, & x > y
%    \end{cases},\\
%    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
%    \mymathbb{0} &= \{\}\infty^{-\infty},\\
%    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0},
%\end{split}
%\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \left({\{\boldsymbol{e}_{i}\}, 1} \right)
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

If one is interested in obtaining only a single MIS configuration, one can just keep a single configuration in the intermediate computations to save the computational effort.
Here is a new algebra defined on the bit strings, replacing the sets of bit strings in \Eq{eq:set}, 
%We leave this as an exercise for readers.
%
%\iffalse
%\begin{equation}
%\begin{split}
%    \sigma_x\infty^x \oplus \sigma_y\infty^y &= \begin{cases}
%        {\rm select}(\sigma_x, \sigma_y)\infty^{\max(x,y)}, & x = y\\
%        \sigma_y\infty^{\max(x,y)}, & x < y\\
%        \sigma_x\infty^{\max(x,y)}, & x > y
%    \end{cases},\\
%    \sigma_x\infty^x \odot \sigma_y\infty^y &= (\sigma_x \lor^\circ \sigma_y)\infty^{x+y},\\
%    \mymathbb{0} &= 1^{\otimes n}\infty^{-\infty},\\
%    \mymathbb{1} &= 0^{\otimes n}\infty^{0},
%\end{split}
%\end{equation}
%\fi
\begin{equation}
\begin{split}
    \sigma \oplus \tau &= {\rm select}(\sigma, \tau), \\
    \sigma \odot \tau &= (\sigma\lor^\circ \tau),\\
    \mymathbb{0} &= 1^{\otimes |V|}, \\
    \mymathbb{1} &= 0^{\otimes |V|},
\end{split}\label{eq:singleconfig}
\end{equation}
where the \texttt{select} function picks one of $\sigma$ and $\tau$ by some criteria to make the algebra commutative and associative, e.g. by picking one with a smaller integer value.
%In practice, we can just pick randomly from them, in which case the program will output one of the MIS configurations randomly.

% Although the above discussion is in the context of the MIS problem, we can also combine the above algebras with the maximal independence network in \Eq{eq:maximal} to enumerate or sample maximal independent sets,
% as well as other problems in \App{app:otherproblems} for configuration enumeration and sampling.

\subsection{Bounding the MIS enumeration space}
When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we use $\alpha(G)$ to bound the searching space.
First, we compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra, \Eq{eq:countingtropicalset}, for obtaining all configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagation of the gradients.
To derive the back-propagation rule for tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ be $\overline C$; the back-propagation rule for gradient masks reads
\begin{equation}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations.
